---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

preview - html: 

```{r}
library(dplyr)

library(gapminder)
data("gapminder")
attach(gapminder)

gapminder <- read.csv("data/gapminder_data.csv", header=TRUE)


head(gapminder)
```
I've loaded the gapminder data and I am going to manipulate it with dplyr.
insert - R, gives a new chunk
```{r}
gap <- gapminder
rm(gapminder)

mean(gap[gap$continent == "Africa", "gdpPercap"])
mean.gdpAfrica <- mean(gap$gdpPercap[continent=="Africa"])
mean.gdpAfrica
```

gap$continent - getting out rows where is Africa

#working with data the dplyr way
select() - subset variables (columns)
filter() - subset observations (rows)
group_by() - counts and groups data
summarize() - useful indormation about the data
mutate() - add a new colum based on the data
```{r}
ncol(gap)

names(gap)

subset_gap <- gap %>% select(year, country, gdpPercap)

subset_gap
dim(subset_gap)


```

filter() - subset on rows
%>% - pipe,

``` {r}
year_country_gdp_euro <- gap %>%
  filter(continent=="Europe") %>%
  select(year, country, gdpPercap)

year_country_gdp_euro

str(year_country_gdp_euro)
```

this is useful for transcriptome data, where want to pull out just some genes from it.

```{r}
year_07 <- gap %>%
  filter(continent=="Africa", year==2007) %>%
  select(year, country, lifeExp)

year_07
```

group_by() - function subset your data by chunking it based on a comlumn
```{r}
head(gap)
str(gap)

gap_by_continent <- gap %>% group_by(continent)
head(gap_by_continent)
str(gap_by_continent)

```
observations (rows) 
variables (columns)
group_by() - turns out a single dataframe into a set of data frames grouped based on some variable

``` {r}
gdp_by_continent <- gap_by_continent %>% 
  summarise(mean_gdpPercap=mean(gdpPercap))

gdp_by_continent

gdp <- gap %>%
  summarise(mean_gdpPercap=mean(gdpPercap))
gdp
```       

#socrative
```{r}

lifeExp_africa <- gap %>%
  group_by(year) %>%
  filter(continent=="Africa") %>%
  group_by(year) %>%
  summarise(mean_lifeExp_africa=mean(lifeExp))
  

lifeExp_africa
```
Calculate the aversge life exp per country. which has de longest average lifexp? and which one has the shortest?
mutate() - function let youadd a column in th data frame based the data
```{r}
life <- gap %>%
  group_by(country) %>%
  summarise(mean_lifeExp = mean(lifeExp))
life

life %>% filter(mean_lifeExp==min(mean_lifeExp))

life %>% filter(mean_lifeExp==max(mean_lifeExp))

```


```{r}
billgdp <- gap %>% 
  filter(year==2007) %>%
  mutate(billion_gdp = gdpPercap * pop / 10^9) %>%
  select (continent, country, billion_gdp)
colnames(billgdp)
dim(billgdp)
```
group more than one thing at a time
```{r}
gdp_by_cont <- gap %>%
  group_by(continent, year) %>%
  summarise(mean_gdp = mean(gdpPercap), 
            sd_gdp = sd(gdpPercap), 
            mean_pop = mean(pop), 
            sample_size = n(), #function n() counts sample size of the variable grouped by
            se_gdp = sd_gdp / sqrt(sample_size))

gdp_by_cont
str(gdp_by_cont)
gdp_by_cont <- gdp_by_cont %>% data.frame()
str(gdp_by_cont)

gdp_by_cont <- data_frame(gdp_by_cont)
```

#combinig dplyr with ggplot2

```{r}
library(ggplot2)

gap %>% filter(continent == "Asia") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

```{r}
gap_wide <- read.csv("data/gapminder_wide.csv", stringsAsFactors = FALSE)

dim(gap_wide)
colnames(gap_wide)
```

Tidy data - table has columns tht are variables and rows as observations. Each of them are single observations. This is the format to use for a scatter plot. 

This data is in a wide format and will change it to a long format
long format - every row has one measurment
To make the data long, use the function gather ()
```{r}
library(tidyr)
head(gap_wide)
gap_long <- gap_wide %>%
  gather(obstype_year, obs_values, -continent, - country)
dim(gap_long)

head(gap_long)
```




```{r}
gap_long <- gap_long %>% separate(obstype_year, into = c("obstype", "year"), sep="_")
gap_long$year <- as.integer(gap_long$year)
head(gap_long)

```





```{r}
mean_LExp <- gap_long %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(lifeExp))

mean_LExp
```




```{r}
gap_long %>% filter(obstype == "lifeExp") %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(obs_values))

```



```{r}
gap_long %>% filter(obstype == "lifeExp") %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(obs_values))

temp_gap <- filter(gap_long, obstype == "lifeExp")
temp_gap <- group_by(temp_gap, continent)
temp_gap <- summarise(temp_gap, mean_lifeExp = mean(obs_values))
temp_gap
```



```{r}
str(gap_long)
gap_normal <- gap_long %>% spread(obstype, obs_values) #spread thevalues along the rows
gap_normal
str(gap_normal)
```

```{r}
str(gap_normal)
colnames(gap_normal)

str(gap)
colnames(gap)
names(gap)
gap_normal <- gap_normal[,names(gap)]

names(gap)
names(gap_normal)
all.equal(gap, gap_normal) #check if these are equal

head(gap)
head(gap_normal)
gap_normal <- gap_normal %>% arrange(country, continent, year)
head(gap_normal)
head(gap)
all.equal(gap, gap_normal)
str(gap)
str(gap_normal)
```



```{r}

```